GMSE: an R package for generalised management strategy evaluation

### 2017-08-07 notebook/ms/ms.Rmd
Change 'stakeholder' to a general term?

Game theory is the standard tool for understanding conflict between rational agents, and is therefore the appropriate tool for modelling manager and stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016]. A game-theoretic approach is compatible with agent-based modelling [@An2012; @Tesfatsion2017], but given the number of possible strategies available to agents, identifying optimal strategies is often intractable. Instead, genetic algorithms, which mimic the process of natural selection (mutation, recombination, selection, reproduction), are useful to find adaptive solutions for game strategies [e.g., @Balmann2000; @Tu2000; @Hamblin2013].

In keeping with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for managers or stakeholders. Instead, within each time step of a simulation (Figure 1), a new genetic algorithm is called for each agent to heuristically find strategies that are well-adapted to achieving agent goals. Hence for the manager and each stakeholder, a unique population with a random set of possible values affecting policy or actions, respectively, is initialised. In each generation of the genetic algorithm, strategies crossover and mutate; when this results in strategies that are over-budget, values are iteratively decreased at random until budget constraints are satisfied. A fitness function then evaluates each strategy in the population, and a tournament is used to select the next generation of strategies [@Hamblin2013]. The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fitness of the fittest strategy between the current and previous generation is below some threshold. The highest fitness strategy in the population then becomes the agent's strategy in the current time step.

@@

Simulations are run using the gmse() function, which offers a range of options for setting parameter values (see Table 1 for some select examples). Output of gmse() is an exhaustive list that includes all resources and observations, all manager and stakeholder decisions and actions, and all landscape properties in each time step of the simulation. Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the plot_gmse_results function). An example below shows how simulations are set and interpreted.

@@

Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of stakeholder land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with farmers [e.g., @Tuvendal2015; @Fox2016a; @Fox2017a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops. Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. Here we consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity (applied to mortality) is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. Managers estimate population size in each time step using mark-recapture techniques, and use their estimates to set costs of culling and scaring (non-lethal) waterfowl (i.e., 'policy') for ten farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

@@

Figure 2 shows the landscape broken down by resource position and stakeholder ownership in the upper left and right hand panels, respectively. The waterfowl population fluctuates around the manager's target size of 1000, but the manager's estimate of population size deviates from its actual size due to uncertainty caused by the observation process (compare black and blue lines the middle left panel). Because the waterfowl have a direct negative effect on landscape yield, total landscape yield (orange line of the middle left panel), along with the yield of individual farmers (right middle panel), is low when waterfowl abundance is high, and vice versa.

@@

GMSE incorporates feedback between resource dynamcs, management policy, and stakeholder actions over time given multiple sources of uncertainty. By simulating complex systems of adaptive management in GMSE, it is possible to gain new insights into questions relevant to conservation. One obvious application of our software is to parameterise simulations with initial conditions similar to those from empirical populations of conservation interest. Running replicate simulations using empirically derived starting conditions could facilitate the prediction of key social-ecological outcomes (e.g., resource extinction, agricultural yield) given uncertainty, and the sensitivity of such outcomes to different management options (e.g., population target, policies available, observation methods, etc.), realised through results from replicate simulations, could inform management decisions.

Additionally, GMSE might be useful for developing adaptive management theory by allowing researchers to explore general questions concerning management *in silico* through simulation. Such questions might include the following: How is population persistence affected by management frequency or observation intensity? How does variation in stakeholder actions affect the distribution of resources or landscape properties? How do asymmetries in manager and stakeholder influence (i.e., budgets) affect resource dynamics? All of these questions can be explored through simulation using GMSE.

@@

The GMSE package is under continued development to include additional features that will be of interest to conservation biologists, managers, and the general public. Such features will include multiple (interacting) resource types and sub-types (e.g., structured populations), improved manager and stakeholder decision-making and prediction based on multiple time steps of simulation history, incorporation of empirical data (e.g., landscape features), and a browser-based graphical user interface. Additionally, future versions of GMSE will allow software users to take the place of managers or stakeholders in decision-making during simulations (replacing the genetic algorithm as desired), facilitating the collection of data to test sociological hypotheses and further improve the realism of the genetic algorithm (for a preliminary example of this, see the 'hunt' argument in the gmse() function). Code underlying GMSE is publicly available on GitHub < https://github.com/bradduthie/gmse > and highly modular, meaning that GMSE model components (black boxes in Figure 1) can be developed freely and independently, then integrated into the broader GMSE framework. The GMSE R package is therefore a versatile and collaborative tool with widespread applications for adaptive resource management.

>>>>

Game theory is the standard tool for understanding conflict between rational agents, and is therefore the appropriate tool for modelling stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016]. A game-theoretic approach is compatible with agent-based modelling [@An2012; @Tesfatsion2017], but given the number of possible strategies available to agents, identifying optimal strategies is often intractable. Instead, genetic algorithms, which mimic the process of natural selection (mutation, recombination, selection, reproduction), are useful to find adaptive solutions for game strategies [e.g., @Balmann2000; @Tu2000; @Hamblin2013].

In keeping with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for stakeholders. Instead, within each time step of a simulation (Figure 1), a new genetic algorithm is called for each stakeholder to heuristically find strategies that are well-adapted to achieving their goals. Hence for the manager and each user, a unique population with a random set of possible values affecting policy or actions, respectively, is initialised. In each generation of the genetic algorithm, strategies crossover and mutate; when this results in strategies that are over-budget, values are iteratively decreased at random until budget constraints are satisfied. A fitness function then evaluates each strategy in the population, and a tournament is used to select the next generation of strategies [@Hamblin2013]. The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fitness of the fittest strategy between the current and previous generation is below some threshold. The highest fitness strategy in the population then becomes the stakeholder's strategy in the current time step.

@@

Simulations are run using the gmse() function, which offers a range of options for setting parameter values (see Table 1 for some select examples). Output of gmse() is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation. Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the plot_gmse_results function). An example below shows how simulations are set and interpreted.

@@

Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of user land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with users [hereafter 'farmers', e.g., @Tuvendal2015; @Fox2016a; @Fox2017a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops. Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. Here we consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity (applied to mortality) is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. Managers estimate population size in each time step using mark-recapture techniques, and use their estimates to set costs of culling and scaring (non-lethal) waterfowl (i.e., 'policy') for ten farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

@@

Figure 2 shows the landscape broken down by resource position and farmer land ownership in the upper left and right hand panels, respectively. The waterfowl population fluctuates around the manager's target size of 1000, but the manager's estimate of population size deviates from its actual size due to observation uncertainty (compare black and blue lines the middle left panel). Because the waterfowl have a direct negative effect on landscape yield, total landscape yield (orange line of the middle left panel), along with the yield of individual farmers (right middle panel), is low when waterfowl abundance is high, and vice versa.

@@

GMSE incorporates feedback between resource dynamcs, management policy, and user actions over time given multiple sources of uncertainty. By simulating complex systems of adaptive management in GMSE, it is possible to gain new insights into questions relevant to conservation. One obvious application of our software is to parameterise simulations with initial conditions similar to those from empirical populations of conservation interest. Running replicate simulations using empirically derived starting conditions could facilitate the prediction of key social-ecological outcomes (e.g., resource extinction, agricultural yield) given uncertainty, and the sensitivity of such outcomes to different management options (e.g., population target, policies available, observation methods, etc.), realised through results from replicate simulations, could inform management decisions.

Additionally, GMSE might be useful for developing adaptive management theory by allowing researchers to explore general questions concerning management *in silico* through simulation. Such questions might include the following: How is population persistence affected by management frequency or observation intensity? How does variation in user actions affect the distribution of resources or landscape properties? How do asymmetries in stakeholder influence (i.e., budgets) affect resource dynamics? All of these questions can be explored through simulation using GMSE.

@@

The GMSE package is under continued development to include additional features that will be of interest to conservation biologists, managers, and the general public. Such features will include multiple (interacting) resource types and sub-types (e.g., structured populations), improved stakeholder decision-making and prediction based on multiple time steps of simulation history, incorporation of empirical data (e.g., landscape features), and a browser-based graphical user interface. Additionally, future versions of GMSE will allow software users to take the place of stakeholders in decision-making during simulations (replacing the genetic algorithm as desired), facilitating the collection of data to test sociological hypotheses and further improve the realism of the genetic algorithm (for a preliminary example of this, see the 'hunt' argument in the gmse() function). Code underlying GMSE is publicly available on GitHub < https://github.com/bradduthie/gmse > and highly modular, meaning that GMSE model components (black boxes in Figure 1) can be developed freely and independently, then integrated into the broader GMSE framework. The GMSE R package is therefore a versatile and collaborative tool with widespread applications for adaptive resource management.
